---
  title: The complete guide to `chouca`
  author: Alexandre Génin <a.a.h.genin@uu.nl>
  output:
    html_document:
      self_contained: no
      toc: yes
      toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.align = "center"
)
```

```{r}
library(chouca)
```

# Introduction

`chouca` is engine in R for 
[stochastic cellular automata](https://en.wikipedia.org/wiki/Stochastic_cellular_automaton), which are models that describes the temporal dynamics of a 2D landscape made of cells. Each cell can be in one of a finite number of states, and switch states over time with 
probabilities that depend on their neighborhood. 

A picture being worth a thousand words, the output of a stochastic cellular automaton 
looks like the following animation in practice: 

```{r gifcreate, echo = FALSE, output = FALSE, fig.show = "animate", fig.width = 3, fig.height = 3, dev='jpeg', interval = 0.04, cache = TRUE}

mod <- ca_library("aridvege") 
init <- generate_initmat(mod, rep(1/3, 3), nr = 64, nc = 64)
export_landscapes <- function(t, mat) { 
  if ( t <= 128 ) {
    return(TRUE)
  }
#   png(sprintf("landscape_%03d.png", t), width = 640, height = 640)
  par(mar = rep(0, 4))
  chouca:::image.camodel_initmat(mat, 
                                 xaxt = "n", yaxt = "n", 
                                 col = hcl.colors(12, "YlOrRd", rev = TRUE))
#   dev.off()
  return(TRUE)
}

invisible({ 
  run_camodel(mod, init, niter = 164, 
            control = list(custom_output_every = 1, 
                           custom_output_fun = export_landscapes, 
                           console_output_every = 0))
})

```

Stochastic cellular automata are widely used to describe the dynamics of landscapes, for 
example, that of forests or coral reefs over space. Yet, implementations are often ad-hoc, 
carried out by the authors themselves, which is prone to bugs and often leads to quite 
slow implementations. `chouca` tries to avoid this by providing a base onto which such 
models can be built. The goal is for the user to declare the probabilities of state 
transitions of the model, then `chouca` handles the rest. 

While `chouca` is built mainly for stochastic cellular automata, it can also handle 
non-stochastic cellular automata, such as the game of life or the [rock-paper-scissor 
automaton](https://www.youtube.com/watch?v=TvZI6Xc0J1Y). If you are specifically 
interested in deterministic cellular automata though, there are probably better 
alternatives to run your model. Chouca supports a wide selection of models, 
while preserving decent performance. This is done by writing most of the code in C++, and 
being able to emit and compile C++ code on the fly for a specific model at run time. This 
enables compiler optimizations that would be impossible with pre-compiled code, and often 
improves simulation speed by one or two orders of magnitude. Chouca can also use a
memoisation-based strategy to run simulations, in which probabilities are not recomputed 
for each cell at each iteration, providing another boost in perfmance. For supported 
models, chouca may probably be 3 to 4 orders of magnitude faster than the common
implementations. 

## Your first model in `chouca`

Let's implement a small forest gap model (Kubo et al. 1996). In such model, cells can be
either 'empty', or filled with a tree, what we call here the 'forest' state. Trees 
produce seeds that can fly over the whole landscape and land haphazardly in any cell. 
Thus, the probability of an empty cell to switch to the forested state is proportional 
to $p_+$, the proportion of cells with trees in the landscape. If we represent the 
empty state as $0$ and the forest state as $+$, this probability of transition is thus: 

$$P(0 \rightarrow +) = \alpha p_{forest}$$ 

where $\alpha$ is the rate at which seeds are produced by each tree. 

Trees naturally die over time, and because wind bursts are stronger in empty areas, trees have a higher probability of dying when they have empty neighboring cells: 

$$P(+ \rightarrow 0) = \delta_0 + \delta q_{empty}$$ 

Here, $q_{empty}$ denotes the proportion of cells around the focal cell that are empty, 
i.e. in the "empty" state. $q_{empty}$ is equal to one when a cell has all of its 
neighbors in the empty state, and zero when none of the neighbors is in this state. 

This model has three parameters, $\alpha$, $\delta_0$ and $\delta$, and can be 
defined in `chouca` using the following syntax: 
```{r example_mod_def}
kubo_model <- camodel(transition(from = "empty",  to = "forest", 
                                 ~ alpha * p["forest"]), 
                      transition(from = "forest", to = "empty",  
                                 ~ delta0 + delta * q["empty"]), 
                      parms = list(alpha = 0.6, delta0 = 0.05, delta = 0.4), 
                      wrap = TRUE, 
                      neighbors = 8)
```

We first declare the two transitions, i.e. the states between which they occur, and an 
expression describing how to compute the probability. In that expression, `p` refers to 
the proportions of cells of the landscape in each state, and `q` refers to the the 
proportion of neighbors in each state. `q["forest"]` for example represents the 
proportion of neighbors in the `forest` state, which is a number between 0 and 1. 

We then set the argument `parms` to a list containing the numerical values of the model
parameters. We use the argument `wrap = TRUE` to state that the model will be run on a 
space in which the edges "wrap around", and the first column/top row of the landscape is neighboring the rightmost/bottom column (a toric space). We use the argument 
`neighbors = 8` to state that we want to consider all 8 neighbors of a cell, 
i.e. including those that are diagonal. This type of neighborhood is referred to as 
[Moore](https://en.wikipedia.org/wiki/Moore_neighborhood) neighborhood, the other 
supported type being a 4-way (Von-Neumann) neighborhood. 

A summary of the model can be printed on the R command line, that recalls the model 
definition and options. 

```{r}
kubo_model
```

The next step is to define the initial landscape from which we want to start the 
simulation. We can do so using the function `generate_initmat`: 

```{r initmat}
init_mat <- generate_initmat(kubo_model, pvec = c(empty = 0.5, forest = 0.5), 
                             nr = 128, nc = 256)
```

Here we initalize the covers to half forest and half empty space using the 
argument `pvec`, and use a landscape of 128 rows by 256 columns. We can now run 
the simulation: 

```{r run}
out <- run_camodel(kubo_model, init_mat, niter = 100)
```

We can then display the results, either in the R console using `summary()`, or as a 
plot using the generic function `plot()`, which displays the global covers of each 
state through time, or `image()`, which displays the landscape: 

```{r display_console}
summary(out)
```

```{r display_plot, fig.width = 12, fig.height = 6}

par(mfrow = c(1, 2))
plot(out)
title("Global covers over time")
image(out)

```



# Supported models 

`chouca` works only with models that use either a 4x4 or 8x8 neighborhood (von Neumann or 
Moore neighborhood, respectively), and in which the effects are isotropic, i.e. all 
neighbors have a similar effect regardless of whether they are above, or below a given 
cell. Cellular automata that involve a preferencial direction cannot be implemented 
in `chouca`, for example with water running downslope in a landscape (Mayor et al. 2013).

The probabilities of transition can have any form, but respecting those two principles 
will work better for `chouca`: 

* To be accurately computed, transition probabilities must be a linear
polynomial of the products of the local or global proportion of cells in a given state. 
For example, a transition probability defined between states $i$ and $j$ as 
$P(i \rightarrow j) = 0.1 + p_i + p_j q_i$ works perfectly, but something involving non-linear functions such as 
$P(0 \rightarrow +) = 0.1 + sin^{p_i}$ may not be able to be correctly handled by 
`chouca` (see below for the consequences). 

* There is some tolerance is discontinuities in the transition rules. Rules such as "if 
one neighbor is in state $i$, then the probability of transition is increased" are OK. 
However, such rule involving multiple types of neighbors are not, e.g. "if one neighbor is 
in state $i$ *and* another one in state $j$, then...". This type of discontinuity is 
hard to accomodate by chouca. 

Overall, the assumed functional form of transitions probabilities must be the following: 

$$ 
P( a \rightarrow b ) = \beta_0 + 
  \sum_{s \in S} f_s(q_s) + 
  \alpha_1 p_1 ... \alpha_n p_S + 
  \beta_1  p_1 q_1, p_1^2 q_1^2 ...  \beta_n p_S^k q_S^k + 
  \gamma_1 q_1 q_2, q_1^2 q_2^2 ... \gamma_n q_(S-1)^k q_S^k + 
  \delta_1 p_1 q_1, p_1^2 q_1^2 ... \delta_n p_S^k q_S^k 
$$

where $S$ represents the set of states of the cellular automaton, $f_s$ can be any 
function of $q_s$. All the $(\alpha)$, $(\beta)$, $(\gamma)$, $(\delta)$ are 
constants (model parameters). Many models can be represented using this functional form - 
however, occasionnaly this may not be the case. This functional form can be daunting, but 
`chouca` will warn you if the transition probabilities cannot be accurately modelled, so 
the best way is to write your model and see if it is supported. 

For all models, `chouca` reports an estimation of the error on the probabilities of 
transition, and whether those can be above one or below zero. For example, the 
above mussel bed model can be perfectly represented by `chouca`, so the reported error 
is zero: 

```{r}
print(kubo_model)
```

The following model, which includes an exponential function of the neighborhood, has a 
non-zero error, because this exponential function cannot be perfectly approximated by 
a polynomial: 

```{r}
mod <- camodel(transition(from = "dead", to = "live", 
                          ~ 0.1 + exp(0.2 * q["dead"] + p["dead"])), 
               transition(from = "live", to = "dead", 
                          ~ 0.1), 
               wrap = TRUE, 
               neighbors = 8)
print(mod)
```

The reported maximum absolute error on the probability of transitions is `r format(max(mod$max_error), digits = 2)`, which is very small. The relative maximum 
error, in other words the error in percentage of the exact probability, is also 
very small (`r format(max(mod$max_rel_error), digits = 2)`). Because both of this 
probabilities are very small, you are probably safe to run this model in chouca and 
obtain correct results. 

The following model gives more trouble: 

```{r}
mod <- camodel(transition(from = "dead", to = "live", 
                          ~ 0.1 + sin(pi * q["dead"] * p["dead"] )), 
               transition(from = "live", to = "dead", 
                          ~ 0.1), 
               wrap = TRUE, 
               neighbors = 8)
print(mod)
```

Here, both the absolute and relative errors on the computed probabilities are 
non-negligible (`r format(max(mod$max_error), digits = 2)` and 
`r format(max(mod$max_rel_error), digits = 2)`). A warning is produced, 
and the problematic transition is pointed out. You should probably revise this model 
if you want to make it work with `chouca`. A possible approach is to approximate 
the function by a taylor expansion of $sin$, for example using the expansion in zero: 
$sin(x) \approx x - \frac{x^3}{3!} + \frac{x^5}{5!}$. 

```{r}

sin_approx <- function(x) { 
  x - x^3/factorial(3) + x^5/factorial(5) 
}

mod <- camodel(transition(from = "dead", to = "live", 
                          ~ 0.1 + sin_approx(pi * q["dead"] * p["dead"]) ), 
               transition(from = "live", to = "dead", 
                          ~ 0.1), 
               wrap = TRUE, 
               neighbors = 8)
print(mod) # negligible error this time, as we use an approximation of sin()

```




# Performance 

One of the main appeal of `chouca` is its performance. Several approaches exist to 
improve simulation speed (i) on-the-fly compilation, (ii) memoisation of transition 
probabilities and (iii) multithreading. 

On-the-fly compilation can be enabled using a control argument to chouca. 

```{r}
control_args <- list(engine = "compiled", 
                     precompute_probas = TRUE, 
                     verbose_compilation = TRUE, 
                     console_output_every = 128) 

out <- run_camodel(kubo_model, 
                   initmat = init_mat, 
                   niter = 512, 
                   control = control_args)
```

On the first run, this model will take a few extra seconds to run as the C++ code 
corresponding to the model needs to be compiled. However, the simulation itself will be 
much faster. Compilation only needs to be done when the structure of the model, or the 
size of the landscape changes. A simple change in parameter value will not trigger a new 
compilation most of the time. 

Another approach to better performance lies in the precomputation of probabilities. The 
classic approach to run a CA is to consider each cell in the landscape one by one, and 
compute its probability of transition to every state it could switch to. This may entail a 
lot of repeated computations, as many cells will have the same neighborhood configuration. 
Instead of doing so, we can compute the probabilities of transition for every neighborhood 
configuration, and just look the number up when considering whether cells switch or not. 
This approach is enabled by using the argument `precompute_probas` in the control list: 

```{r}

control_args <- list(engine = "compiled", 
                     precompute_probas = TRUE, 
                     console_output_every = 128) 

out <- run_camodel(kubo_model, 
                   initmat = init_mat, 
                   niter = 512, 
                   control = control_args)

```

This approach can dramatically improve the performance for models that have a 
small number of states and simple transition rules. However, it may be 
counter-productive for models with a larger number of states, as the number of 
neighborhood configurations grows very quickly with the number of different model 
states ($S^4$ or $S^8$ depending on the type of neighborhood). By default, a simple
heuristic is used to make a decent choice. 

The last approach to improve performance is to use multithreading. In chouca, the 
approach is to consider the updating of cells in parallel, instead of doing it 
serially. However, there is some synchronisation work required between threads to make 
sure the different cores are not writing to shared data structures. This often results
in a significant overhead, and only small performance improvements. 

```{r}

control_args <- list(engine = "compiled", 
                     precompute_probas = TRUE, 
                     cores = 4, 
                     console_output_every = 128) 

out <- run_camodel(kubo_model, 
                   initmat = init_mat, 
                   niter = 512, 
                   control = control_args)

```

Future improvements of the package may improve this situation. In the meantime, if you 
happen to be running several simulations at the same time (which is often the case), you 
are probably better off parallelizing at a higher level. 

Finally, because the model code is compiled on the fly, you may be able to use 
optimizations that are specific to your combination of platform and compiler. The best 
way to do so is to edit your platform '~/.R/Makevars' file to add the flags you want, 
such as `CXXFLAGS=-O3 -march=native -mtune=native'. However, do not expect too much as 
most of the time this does not bring much more performance to the table. 


# Storing/discarding data 



# Visualising simulations 

`chouca` can display the covers of each state, or the complete landscape as the 
simulation is running. 

# Warning about error in probabilites 


When writing a transition probability, `chouca` will try to approximate it with the above 
function, check that how accurately this has been done, and report any deviation. For 
example, the following model has transition probabilities that cannot be represented with 
100% accuracy, because of the exponential term, and as a result a warning is produced: 

```{r model_error, results = FALSE}
camodel(transition(from = "0", to = "+", ~ 1 + exp( 1 + - p["+"] * q["+"])), 
        wrap = TRUE, 
        neighbors = 4)
```

```{r model_error_hidden, echo = FALSE, results = TRUE, warning = FALSE}
invisible({ 
  mod <- camodel(transition(from = "0", to = "+", ~ 1 + exp(- p["+"] * q["+"])), 
                 wrap = TRUE, 
                 neighbors = 4)
  relerr <- max(mod$max_rel_error)
})
```

`chouca` reports both the absolute and relative difference between the approximated and 
true model probability. Many models have no error and no approximation is made, but you 
can go on with a model with error if you know what you are doing. Here, the relative error 
is tiny, so the results would probably be extremely close to a model without 
approximation. 

To get rid of this warning, we can provide our own approximation to make the above 
form into something more palatable to `chouca`, by using the first few terms of the 
(Maclaurin)[https://en.wikipedia.org/wiki/Taylor_series#Exponential_function] series for 
the exponential function: $e^x = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!}$. 

```{r model_error_ok}
exp_approx <- function(x) 1 + x + x^2/(2*1) + x^3/(3*2*1) + x^4/(4*3*2)
camodel(transition(from = "0", to = "+", ~ 1 + exp_approx(- p["+"] * q["+"])), 
        wrap = TRUE, 
        neighbors = 4)
```

This time, no error is reported by chouca, as this new model follows the exact functional 
form required. 



# References 

Kubo, Takuya, Yoh Iwasa, and Naoki Furumoto. 1996. “Forest Spatial Dynamics with Gap Expansion: Total Gap Area and Gap Size Distribution.” Journal of Theoretical Biology 180 (3): 229–46.

Mayor, Ángeles G., Sonia Kéfi, Susana Bautista, Francisco Rodríguez, Fabrizio Cartení, and Max Rietkerk. 2013. “Feedbacks between Vegetation Pattern and Resource Loss Dramatically Decrease Ecosystem Resilience and Restoration Potential in a Simple Dryland Model.” Landscape Ecology 28 (5): 931–42. https://doi.org/10.1007/s10980-013-9870-4.
