---
  title: The complete guide to `chouca`
  author: Alexandre Génin <a.a.h.genin@uu.nl>
  output:
    html_document:
      self_contained: no
      toc: yes
      toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.align = "center"
)
```

# Introduction

`chouca` is engine in R for 
[stochastic cellular automata](https://en.wikipedia.org/wiki/Stochastic_cellular_automaton), which are models that describes the temporal dynamics of a 2D landscape made of cells. Each cell can be in one of a finite number of states. In such model, time is discrete, and at each iteration cells have a probability of switching from one state to another that typically depends on model parameters, and the state of their neighbors. 

A picture being worth a thousand words, the output of a stochastic cellular automaton 
looks like the following animation in practice: 

```{r gifcreate, echo = FALSE, output = FALSE, fig.show = "animate", fig.width = 3, fig.height = 3, dev='jpeg', interval = 0.04, cache = TRUE}

mod <- ca_library("aridvege") 
init <- generate_initmat(mod, rep(1/3, 3), nr = 64, nc = 64)
export_landscapes <- function(t, mat) { 
  if ( t <= 128 ) {
    return(TRUE)
  }
#   png(sprintf("landscape_%03d.png", t), width = 640, height = 640)
  par(mar = rep(0, 4))
  chouca:::image.camodel_initmat(mat, 
                                 xaxt = "n", yaxt = "n", 
                                 col = hcl.colors(12, "YlOrRd", rev = TRUE))
#   dev.off()
  return(TRUE)
}

invisible({ 
  run_camodel(mod, init, niter = 164, 
            control = list(custom_output_every = 1, 
                           custom_output_fun = export_landscapes, 
                           console_output_every = 0))
})

```

Stochastic cellular automata are widely used to describe the dynamics of landscapes, for 
example, that of forests or coral reefs over space. Yet, implementations are often ad-hoc, 
carried out by the authors themselves, which is prone to bugs and often leads to quite 
slow implementations. `chouca` tries to avoid this by providing a base onto which such 
models can be built. The goal is for the user to declare the probabilities of state 
transitions of the model, then `chouca` handles the rest. 

While `chouca` is built mainly for stochastic cellular automata, it can also handle 
non-stochastic cellular automata, such as the game of life or the [rock-paper-scissor 
automaton](https://www.youtube.com/watch?v=TvZI6Xc0J1Y). If you are specifically 
interested in deterministic cellular automata though, there are probably better 
alternatives to run your model. Chouca supports a wide selection of cellular automata, 
while preserving decent performance. This is done by writing most of the code in C++, and 
being able to emit and compile C++ code on the fly for a specific model at run time. This 
enables compiler optimizations that would be impossible with pre-compiled code, and often 
improves simulation speed 10-100x. Chouca can also use a memoisation-based strategy to 
run simulations, in which probabilities are not recomputed for each cell at each 
iteration, providing another boost in perfmance. Overall, chouca may probably be 3 to 
4 orders of magnitude faster than the average model implementation found in the 
literature. 

## A small example

Let's define a small forest model. In such model, cells can be either 'empty', or filled 
with a tree, what we call here the 'forest' state. Trees produce seeds that can fly over 
the whole landscape and land haphazardly in any cell. Thus, the probability of an empty 
cell to switch to the forested state is proportional to $p_+$, the proportion of cells 
with trees in the landscape. If we represent the empty state as $0$ and the forest state 
as $+$, this probability of transition is thus: 

$$P(0 \rightarrow +) = \alpha p_+$$ 

where $\alpha$ is the rate at which seeds are produced by each tree. 

Trees naturally die over time, and because wind bursts are stronger in empty areas, trees have a higher of dying when they have empty neighboring cells: 

$$P(+ \rightarrow 0) = \delta_0 + \delta q_{0}$$ 

Here, $q_0$ denotes the proportion of cells around the focal cell that are empty, i.e. in 
the $0$ state. 

This model has three parameters, $\alpha$, $\delta_0$ and $\delta$, and can be 
defined in `chouca` using the following syntax: 
```{r example_mod_def}
model <- camodel(transition(from = "0", to = "+", ~ alpha * p["+"]), 
                 transition(from = "+", to = "0", ~ delta0 + delta * q["0"]), 
                 parms = list(alpha = 0.6, delta0 = 0.05, delta = 0.4), 
                 wrap = TRUE, 
                 neighbors = 8)
```

We first declare the two transitions, i.e. the states between which it occurs, and an 
expression describing how to compute the probability. In that expression, `p` refers to 
the proportions of cells of the landscape in each state, and `q` refers to the the 
proportion of neighbors in each state. `q["+"]` for example represents the proportion 
of neighbors in the `+` (forest) state. We then set the argument `parms` to a list 
containing the numerical values of the model parameters. We use the argument `wrap = TRUE` 
to state that the model will be run on a space in which the edges "wrap around", and the 
first column/top row of the landscape is neighboring the rightmost/bottom column. We use 
the argument `neighbors = 8` to state that we want to consider all 8 neighbors of a cell, 
i.e. including those that are diagonal. This type of neighborhood is referred to as 
[Moore](https://en.wikipedia.org/wiki/Moore_neighborhood) neighborhood, the other 
type of neighborhood being a 4-way (Von-Neumann) neighborhood. 

A summary of the model can be printed on the R command line, that recalls the model 
parameters, along with the 

```{r}
model
```

The next step is to define the initial landscape from which we want to start the 
simulation. We can do so using the function `generate_initmat`: 

```{r initmat}
init_mat <- generate_initmat(model, pvec = c(`0` = 0.5, `+` = 0.5), nr = 128, nc = 256)
```

Here we initalize the covers to half tree (`+`) and half empty space using the 
argument `pvec`, and use a landscape of 128 rows by 256 columns. We can now run 
the simulation: 

```{r run}
out <- run_camodel(model, init_mat, niter = 100)
```

We can then display the results, either in the R console, or as a plot: 

```{r display_console}
summary(out)
```

```{r display_plot, fig.width = 12, fig.height = 6}

par(mfrow = c(1, 2))
plot(out)
title("Global covers over time")
image(out)

```



# Supported models 

`chouca` works only with models that use either a 4x4 or 8x8 neighborhood (von Neumann or 
Moore neighborhood, respectively), and in which the effects are isotropic, i.e. all 
neighbors have a similar effect regardless of whether they are above, or below a given 
cell. 

The probabilities of transition can have any form, but respecting those two principles will work better for `chouca`: 

* To be 100% accurate, transition probabilities must be a linear combination of the 
products of the local or global proportion of cells in a given state. For example, 
a transition probability defined as $P(0 \rightarrow +) = 0.1 + p_+ + p_0 q_0$ works perfectly, but something involving non-linear functions such as 
$P(0 \rightarrow +) = 0.1 + e^{- p_+}$ will not be ideal (see below for the 
consequences). 

* The transition probabilities can be any function of the proportion of neighbors in a 
given state. Rules such as "if one neighbor is in state $0$, then the probability of 
transition is modified" are OK. Such rules involving multiple types of neighbors are not, 
e.g. "if one neighbor is in state $0$ *and* another one in state $+$, then...". 

Overall, the functional form of transitions probabilities must be the following: 

$$ 
P( a \rightarrow b ) = \beta_0 + 
  \sum_{s \in S} f_s(q_s) + 
  \sum_{s \in S, t \in S} \beta_{qq,s,t} q_s^{ \alpha_{qq, s} } q_t^{ \alpha_{qq, t} } + 
  \sum_{s \in S, t \in S} \beta_{pq,s,t} p_s^{ \alpha_{pq, s} } q_t^{ \alpha_{pq, t} } + 
  \sum_{s \in S, t \in S} \beta_{pp,s,t} p_s^{ \alpha_{pp, s} } p_t^{ \alpha_{pp, t} } 
$$

where $S$ represents the set of states of the cellular automaton, $f_s$ can be any 
function, the $\beta_{x,y,z}$s are constants, and the $\alpha_{x,y}$ are integer 
constants. 

When writing a transition probability, `chouca` will try to approximate it with the above 
function, check that how accurately this has been done, and report any deviation. For 
example, the following model has transition probabilities that cannot be represented with 
100% accuracy, because of the exponential term, and as a result a warning is produced: 

```{r model_error, results = FALSE}
camodel(transition(from = "0", to = "+", ~ 1 + exp( 1 + - p["+"] * q["+"])), 
        wrap = TRUE, 
        neighbors = 4)
```

```{r model_error_hidden, echo = FALSE, results = TRUE, warning = FALSE}
invisible({ 
  mod <- camodel(transition(from = "0", to = "+", ~ 1 + exp(- p["+"] * q["+"])), 
                 wrap = TRUE, 
                 neighbors = 4)
  relerr <- max(mod$max_rel_error)
})
```

`chouca` reports both the absolute and relative difference between the approximated and 
true model probability. Many models have no error and no approximation is made, but you 
can go on with a model with error if you know what you are doing. Here, the relative error 
is tiny, so the results would probably be extremely close to a model without 
approximation. 

To get rid of this warning, we can provide our own approximation to make the above 
form into something more palatable to `chouca`, by using the first few terms of the 
(Maclaurin)[https://en.wikipedia.org/wiki/Taylor_series#Exponential_function] series for 
the exponential function: $e^x = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!}$. 

```{r model_error_ok}
exp_approx <- function(x) 1 + x + x^2/(2*1) + x^3/(3*2*1) + x^4/(4*3*2)
camodel(transition(from = "0", to = "+", ~ 1 + exp_approx(- p["+"] * q["+"])), 
        wrap = TRUE, 
        neighbors = 4)
```

This time, no error is reported by chouca, as this new model follows the exact functional 
form required. 

# `chouca` goals: 

* `chouca` is fast, although some things could be faster. It depends on your model complexity. 

* `chouca` wants to be fast, and help you explore your model behavior. 



## Supported types of models 

`chouca` does not support all kinds of cellular automata. Probabilities must follow the 
following form. Calling x, y z, 

If a model does not follow this, the probabilities of transition will be approximate. The
good news is that `chouca` is able to tell you the error, both relative and absolute: it 
is up to you to decide whether this is a sufficiently good approximation. 




# Visualising simulations 

`chouca` can display the covers of each state, or the complete landscape as the 
simulation is running. 

# Performance 

Parallel processing

# Examples 

## A first example 

## A more complex example 

## A model with errors 

